PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-10-27 IEEEtran/paper.tex
Fix Tristan's comments

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing for simulating
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
In our experiments, we used MCA in Random Rounding (RR) mode, which
simulates roundoff errors by applying function $inexact$ only to the
output returned by a function. The Precision Bounding (PB) mode 
\add{that applies the $inexact$ function on inputs}
can identify other types of instabilities, in particular catastrophic
cancellations, however, it is more difficult to apply to large code bases
as it often leads to execution errors that require human intervention.

@@

        different control flow paths (parsing). \textit{nan} stands for (\texttt{Not-A-Number}). 

@@

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on SciPy
and scikit-learn examples. Columns \textit{trace} and \textit{parse} report on
the tracing and aggregation steps with two possible outcomes: green
\valid\xspace for success and red \cross\xspace for error. Tracing errors occur
in the application as a result of numerical perturbations, whereas parsing
errors occur when traces result from different control flow paths (CFP). 
\add{Hence, an application that consistently fails at the same instruction will
    produce traces with equivalent CFPs which can be parsed by \pytracer as the 
    Face Recognition shows.}

@@

Figure~\ref{fig:spline2d_rr} shows the results of the spline evaluation with
\texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
    [-1,1]\times[-1,1]$. The result is a 2D array that is natively supported by
PyTracer. We observe a significant loss of result precision in some regions, which follows
an interesting pattern that could be further investigated.
\add{This example shows that \pytracer works on 2D inputs and the importance of
    visualizing complex data and not only reporting the values.}

@@

\pytracer's tracing overhead consists of (1) module instrumentation time
(initialization overhead) and (2) data write time during the execution (runtime
overhead). The initialization overhead is proportional to the number of
functions and classes traced. For example, we measured an average initialization
time of 8 seconds during our experiments with  NumPy, SciPy, and Scikit-learn.
This overhead can be reduced by excluding intensively used functions such as
\texttt{numpy.array} from the instrumentation.
Figure~\ref{fig:performance_tracing} summarizes \pytracer's tracing overhead for
SciPy and scikit-learn examples. \textit{Instrumentation and initialization} represents the slowdown of
\pytracer without fuzzy, \add{measured by applying PyTracer to Python programs without Monte-Carlo Arithmetic.}
The slowdown is measured as the ratio between the actual execution time
(without \pytracer) and the execution time using \pytracer. 
The \textit{Total with RR MCA} shows the slowdown within the Fuzzy environment.
Without Fuzzy, \pytracer introduces an average slowdown of $17\times$ for
the examples running less than 8 seconds because \pytracer spends most of its
time in the initialization step. Since this step is a constant factor of the
number of functions instrumented, we also measured the instrumentation cost
itself (\textit{Instrumentation only}) calculated by subtracting the
initialization time to \textit{Instrumentation}. We then obtain an average slowdown of
$3.5\times$. Finally, using MCA
operations (through Fuzzy) instead of the native ones increased the average
slowdown to $37\times$.
\add{As shown in Figure~\ref{fig:performance_parsing}, the log aggregation time is linearly correlated with the log trace size (\todo{$R^2$=}, p-value $\leq 10^{-8}$, \todo{Add test name}).}

>>>>

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing for simulating
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi\! \in\! (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
In our experiments, we used MCA in Random Rounding (RR) mode, which
simulates roundoff errors by applying function $inexact$ only to the
output returned by a function. The Precision Bounding (PB) mode 
\add{that applies the $inexact$ function on inputs}
can identify other types of instabilities, in particular catastrophic
cancellations, however, it is more difficult to apply to large code bases
as it often leads to execution errors that require human intervention.

@@

        different control flow paths (parsing). \textit{nan} stands for (\texttt{Not-A-Number})
        \add{meaning that the output result comes from an invalid arithmetic operation}. 

@@

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on SciPy
and scikit-learn examples. Columns \textit{trace} and \textit{parse} report on
the tracing and aggregation steps with two possible outcomes: green
\valid\xspace for success and red \cross\xspace for error. Tracing errors occur
in the application as a result of numerical perturbations, whereas parsing
errors occur when traces result from different control flow paths (CFP). 
\add{Therefore, an application that consistently fails at the same function call will
  produce traces with equivalent Control Flow Paths which can be parsed by \pytracer.
  In our experiments, this was the case for the Face Recognition shows.}

@@

Figure~\ref{fig:spline2d_rr} shows the results of the spline evaluation with
\texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
    [-1,1]\times[-1,1]$. The result is a 2D array that is natively supported by
PyTracer. We observe a significant loss of result precision in some regions, which follows
an interesting pattern that could be further investigated.
\add{This example shows that \pytracer works on 2D inputs and the importance of
    visualizing complex data types in addition to summary statistics.}

@@

\pytracer's tracing overhead consists of (1) module instrumentation time
(initialization overhead) and (2) data write time during the execution (runtime
overhead). The initialization overhead is proportional to the number of
functions and classes traced. For example, we measured an average initialization
time of 8 seconds during our experiments with  NumPy, SciPy, and Scikit-learn.
This overhead can be reduced by excluding intensively used functions such as
\texttt{numpy.array} from the instrumentation.
Figure~\ref{fig:performance_tracing} summarizes \pytracer's tracing overhead for
SciPy and scikit-learn examples. \textit{Instrumentation and initialization} represents the slowdown of
\pytracer without fuzzy, \add{measured by applying PyTracer to Python programs without Monte-Carlo Arithmetic.}
The slowdown is measured as the ratio between the actual execution time
(without \pytracer) and the execution time using \pytracer. 
The \textit{Total with RR MCA} shows the slowdown within the Fuzzy environment.
Without Fuzzy, \pytracer introduces an average slowdown of $17\times$ for
the examples running less than 8 seconds because \pytracer spends most of its
time in the initialization step. Since this step is a constant factor of the
number of functions instrumented, we also measured the instrumentation cost
itself (\textit{Instrumentation only}) calculated by subtracting the
initialization time to \textit{Instrumentation}. We then obtain an average slowdown of
$3.5\times$. Finally, using MCA
operations (through Fuzzy) instead of the native ones increased the average
slowdown to $37\times$.
\add{As shown in Figure~\ref{fig:performance_parsing}, the log aggregation time is linearly correlated with the log trace size (F-score=68, p-value $\leq 10^{-8}$, obtained with the F-test for linear regression).}
